Polyester RNA-seq simulation study

### 2018-04-10 datasets/RNA-Seq/polyester-simulation.Rmd
added covariateLinePlot() functions and converted to relative paths

# comparison tools/functions
library(SummarizedBenchmark)

# load helper functions
working_dir <- list(project_home = "/users/shicks1/projects/benchmark-fdr")
sourceDirectory(file.path(working_dir$project_home,"datasets/R/"))

@@

  # built Benchmark data
  sb <- bd %>% buildBench(data=geneExp, parallel = FALSE, 
                          BPPARAM = BPPARAM, truthCols = "qvalue", 
                          ftCols = "ind_covariate")
  rowData(sb)$log2FC <- geneExp$effect_size

@@

```{r}
B <- 10
bd <- initializeBenchDesign() # only needs to be done once
excludeSet <- c("unadjusted", "bl-df02", "bl-df04", "bl-df05")
ualpha <- 0.05
```

@@

```{r}
#' @param res standardized metric data.table generated using
#'        standardize_results.
#' @param alpha alpha cutoff
#' @param filterSet which methods to exclude from consideration 
numberMethodsReject <- function(res, alphacutoff, filterSet){
  res <- res %>% 
    filter(is.na(param.alpha) | (param.alpha == alphacutoff)) %>%
    filter(!(blabel %in% filterSet)) %>%
    filter(alpha == alphacutoff) %>%
    filter(performanceMetric == "rejections") %>%
    select(blabel, performanceMetric, value) %>%
    group_by(blabel) %>%
    summarize(mean_value = mean(value)) %>%
    filter(mean_value > 0)
  return(nrow(res))
}

addDefaultMetrics <- function(sb) {
    sb <- addPerformanceMetric(sb, evalMetric = c("TPR", "FPR", "TNR", "FNR", "rejections"),
                               assay = "qvalue")
    return(sb)
}

@@

if (!file.exists(file.path(working_dir, "datasets/RNA-Seq/polyester-results-null5.RDS"))){
  null5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(null5, file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-null5.RDS"))
} else {
  null5 <- readRDS(file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-null5.RDS"))
}
```

@@

res5 <- plotsim_standardize(null5, alpha = seq(0.01, 0.20, 0.01))

@@

if (numberMethodsReject(res5, alphacutoff=ualpha, filterSet=excludeSet) >= 3){
  aggupset(null5, alpha=ualpha, supplementary = FALSE, return_list = FALSE) 
} else {
  message("Not enough methods found rejections at alpha ", ualpha, 
          "; skipping upset plot")
}
```

@@

if (!file.exists(file.path(working_dir, "datasets/RNA-Seq/polyester-results-null10.RDS"))){
  null10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(null10, file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-null10.RDS"))
} else {
  null10 <- readRDS(file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-null10.RDS"))
}
```

@@

```{r}
res10 <- plotsim_standardize(null10, alpha = seq(0.01, 0.20, 0.01))

# metrics = ""TPR"   "FPR"   "TNR"   "FNR"   "rejections"  "FWER"   "rejectprop"
plotsim_average(res10, met="rejections",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res10, met="TNR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

@@

if (!file.exists(file.path(working_dir, "datasets/RNA-Seq/polyester-results-de5.RDS"))){
  de5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(de5, file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-de5.RDS"))
} else {
  de5 <- readRDS(file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-de5.RDS"))
}
```

@@

```{r}
plotEffectSize <- function(sbl, alpha=0.05, DE=TRUE, covname){

summarize_one_item <- function(object, alpha){
  object <- object[,!( grepl("^ihw", as.character( colData( object )$blabel ) )
                       & colData( object )$param.alpha != alpha )]

  df <- as.data.frame(cbind(as.matrix(rowData(object)), 
                            1*(assays(object)[["qvalue"]]) < 0.05))
  colnames(df)[1] <- "truth" 
  df <- df %>%
    select(truth, covname, bonf, bh,
                  qvalue, contains("ihw"), ashs, "bl-df03", lfdr, 
                  "scott-theoretical", "scott-empirical") %>%
    # rename("scott-theoretical"="scott-t", "scott-empirical"="scott-e") %>%
    gather(method, significant, -truth, -covname) %>%
    group_by(method, truth, significant) %>%
    summarize(mean_effect_size=mean(abs(get(covname))),
              med_effect_size=median(abs(get(covname))),
              min_effect_size=min(abs(get(covname))),
              max_effect_size=max(abs(get(covname))),
              botQ=quantile(abs(get(covname)),0.25),
              topQ=quantile(abs(get(covname)), 0.75)) %>% na.omit()
  df$significant <- as.factor(df$significant)
  levels(df$significant)=c("Not Significant", "Significant")
  df$truth <- as.factor(df$truth)
  levels(df$truth)=c("Not DE", "DE")
  return(df)
}

df <- lapply(sbl, summarize_one_item, alpha=alpha)
df <- bind_rows(df, .id = "rep")
df <- as.tibble(df)

if (DE){
  df <- df %>% filter(truth=="DE")
}else{
  df <- df %>% filter(truth=="Not DE")
}

p <- ggplot(df, aes(x = as.factor(method), y=med_effect_size)) +
  geom_boxplot() +
  facet_grid(significant~., scales="free_y") +
  xlab("Method") +
  ylab(paste0("Median ", covname))

if (DE){
  p <- p + ggtitle(paste0(covname, " of True DE genes by significance at alpha ", 
                       alpha))
}else{
  p <- p + ggtitle(paste0(covname, " of Non-DE genes by significance at alpha ", 
                       alpha))
}

return(p)
}

@@

# metrics = ""TPR"   "FPR"   "TNR"   "FNR"   "rejections"  "FWER"   "rejectprop"
plotsim_average(res5d, met="rejections",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res5d, met="FPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res5d, met="TPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

plotEffectSize(de5, covname="log2FC")
plotEffectSize(de5, covname="log2FC", DE=FALSE)

plotEffectSize(de5, covname="ind_covariate")
plotEffectSize(de5, covname="ind_covariate", DE=FALSE)

@@

if (!file.exists(file.path(working_dir, "datasets/RNA-Seq/polyester-results-de10.RDS"))){
  de10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(de10, file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-de10.RDS"))
} else {
  de10 <- readRDS(file=file.path(working_dir, "datasets/RNA-Seq/polyester-results-de10.RDS"))
}
```

@@

# metrics = ""TPR"   "FPR"   "TNR"   "FNR"   "rejections"  "FWER"   "rejectprop"
plotsim_average(res10d, met="rejections",filter_set = excludeSet, 
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res10d, met="FPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res10d, met="TPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

plotEffectSize(de10, covname="log2FC")
plotEffectSize(de10, covname="log2FC", DE=FALSE)

plotEffectSize(de10, covname="ind_covariate")
plotEffectSize(de10, covname="ind_covariate", DE=FALSE)

>>>>

# comparison tools/functions
# devtools::install_github("areyesq89/SummarizedBenchmark", ref="fdrbenchmark") # needed b/c BioC devel branch is 
library(SummarizedBenchmark)

# load helper functions
sourceDirectory(file.path("../R/"))

@@

  # built Benchmark data
  sb <- bd %>% buildBench(data=geneExp, parallel = FALSE, 
                          truthCols = "qvalue", 
                          ftCols = "ind_covariate")
  rowData(sb)$log2FC <- geneExp$effect_size

@@

```{r}
B <- 100
bd <- initializeBenchDesign() # only needs to be done once
excludeSet <- c("unadjusted", "bl-df02", "bl-df04", "bl-df05") 
ualpha <- 0.05
```

@@

```{r}
#' @param res standardized metric data.table generated using
#'        standardize_results.
#' @param alpha alpha cutoff
#' @param filterSet which methods to exclude from consideration 
numberMethodsReject <- function(res, alphacutoff, filterSet){
  res <- res %>% 
    filter(is.na(param.alpha) | (param.alpha == alphacutoff)) %>%
    filter(!(blabel %in% filterSet)) %>%
    filter(alpha == alphacutoff) %>%
    filter(performanceMetric == "rejections") %>%
    select(blabel, performanceMetric, value) %>%
    group_by(blabel) %>%
    summarize(mean_value = mean(value)) %>%
    filter(mean_value > 0)
  return(nrow(res))
}
```

@@

if (!file.exists(file.path("polyester-results-null5.RDS"))){
  null5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(null5, file=file.path("polyester-results-null5.RDS"))
} else {
  null5 <- readRDS(file=file.path("polyester-results-null5.RDS"))
}
```

@@

```{r}
res5 <- plotsim_standardize(null5, alpha = seq(0.01, 0.10, 0.01))

@@

covariateLinePlot(null5, alpha=0.05, covname="log2FC", nbins=25, 
                 trans="log1p")
covariateLinePlot(null5, alpha=0.05, covname="ind_covariate", nbins=25, 
                 trans="log1p")

if (numberMethodsReject(res5, alphacutoff=ualpha, filterSet=excludeSet) >= 3){
  aggupset(null5, alpha=ualpha, supplementary = FALSE, return_list = FALSE) 
} else {
  message("Not enough methods found rejections at alpha ", ualpha, 
          "; skipping upset plot")
}

@@

if (!file.exists(file.path("polyester-results-null10.RDS"))){
  null10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(null10, file=file.path("polyester-results-null10.RDS"))
} else {
  null10 <- readRDS(file=file.path("polyester-results-null10.RDS"))
}
```

@@

```{r}
res10 <- plotsim_standardize(null10, alpha = seq(0.01, 0.10, 0.01))

plotsim_average(res10, met="rejections",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res10, met="TNR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

covariateLinePlot(null10, alpha=0.05, covname="log2FC", nbins=25, 
                 trans="log1p")
covariateLinePlot(null10, alpha=0.05, covname="ind_covariate", nbins=25, 
                 trans="log1p")

@@

if (!file.exists(file.path("polyester-results-de5.RDS"))){
  de5 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(de5, file=file.path("polyester-results-de5.RDS"))
} else {
  de5 <- readRDS(file=file.path("polyester-results-de5.RDS"))
}
```

@@

@@

plotsim_average(res5d, met="rejections",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res5d, met="FPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res5d, met="TPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

covariateLinePlot(de5, alpha=0.05, covname="log2FC", nbins=25, 
                 trans="log1p")
covariateLinePlot(de5, alpha=0.05, covname="ind_covariate", nbins=25, 
                 trans="log1p")

@@

if (!file.exists(file.path("polyester-results-de10.RDS"))){
  de10 <- bplapply(X=1:B, FUN=simulateOneSplit, counts=counts,
                   paramPolyester=params, nDE=nDE, sampleSize=sampleSize, bd=bd,
                   BPPARAM=multicoreParam)
  saveRDS(de10, file=file.path("polyester-results-de10.RDS"))
} else {
  de10 <- readRDS(file=file.path("polyester-results-de10.RDS"))
}
```

@@

plotsim_average(res10d, met="rejections",filter_set = excludeSet, 
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res10d, met="FPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE) 
plotsim_average(res10d, met="TPR",filter_set = excludeSet,
                merge_ihw = TRUE, errorBars=TRUE)

covariateLinePlot(de10, alpha=0.05, covname="log2FC", nbins=25, 
                 trans="log1p")
covariateLinePlot(de10, alpha=0.05, covname="ind_covariate", nbins=25, 
                 trans="log1p")
